---
title: "Using EpiNOAA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{using-epinoaa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
options(download.file.method = "wget")
library(tidyverse)
library(EpiNOAA)
library(viridis)
```

# Near Real Time vs Historical Data

The EpiNOAA package can pull both near-real time preliminary data and historical scaled data.  By default, the EpiNOAA package will merge the two together seamlessly to create a complete timeseries.  This might not be recommended for your use case, however, so please consult the original nClimGrid Documentation.  In both cases, the data are generated through an event-driven pipeline on AWS that runs in a few seconds after receiving new data from NOAA.  

You can filter for scaled and prelim data as we will explore below:


## Near Real Time Data

Let's pull in some near real time data up until today.  This vignette is being written in June of 2023.  Let's pull in data from May and June and take a look at average temperature for Buncombe County, NC. 



```{r prelim-data-cty}
recent_data <- read_nclimgrid_epinoaa(
  beginning_date = '2023-05-01',
  end_date = '2023-06-16',
  spatial_res = 'cty',
  states = c('NC'),
  counties = c('NC: Buncombe County')
  )
```

Terrific!  Now, let's plot the time series of average temperature over the past month and a half highlighting the scaled and prelim values:

```{r recent-ts}

recent_data %>% 
  mutate(tavg = na_if(as.numeric(tavg), -999.99)) %>%
  ggplot(aes(x = date, y = tavg, color = STATUS)) + 
  geom_line() + 
  theme_bw(12) + 
  labs(x = 'Average Daily Temperature', y = '')

```


## Near Real Time County Exploration


In continuing our exploration of recent county data, let's find the coldest counties on the most recent day available.

```{r prelim-data}
county_data <- read_nclimgrid_epinoaa(
  beginning_date = '2023-06-01',
  end_date = '2023-06-15',
  spatial_res = 'cty',
  )
```

```{r recent-day}
county_data %>% 
  filter(date == max(date)) %>% 
  pull(date) %>% 
  unique()
```


find out the most recent day available from our data: 

Let's find the coldest counties in the continental US on that day...

```{r coldest-day,  fig.width=6, fig.height=5}
coldest_counties <- county_data %>% 
  filter(date == max(date)) %>% 
  mutate(tmin = na_if(as.numeric(tmin), -999.99)) %>% 
  arrange(tmin) %>% 
  head(10)

ggplot(coldest_counties, aes(x = region_name, y = tmin)) + 
  geom_bar(stat = 'identity') + 
  theme_bw(14) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  labs(x = '', y = 'Minimum Daily Temperature')
```


# State Data

Let's explore downloading state data for the US for the year of 2021.  We'll look at which state had the highest and lowest temperature during that time and plot average temperature over the course of the year for each of those two states.  

First, let's import the data:
```{r state-import}

state_2021 <-  read_nclimgrid_epinoaa(
  beginning_date = '2021-01-01', 
  end_date = '2021-12-31', 
  spatial_res = 'ste')

```


Let's take a look at the data:
```{r state_examine}
glimpse(state_2021)
```

Note that everything is read in as characters.  We will change that in the next step, then find the state with the highest and lowest temperature:
```{r state-maxmin}
max_min_temp <- state_2021 %>%
  mutate(tmax = as.numeric(tmax),
         tmin = as.numeric(tmin)) %>% 
  group_by(state_name) %>% 
  summarize(max_temp = max(tmax),
            min_temp = min(tmin))

max_temp_state <- max_min_temp %>% 
  filter(max_temp == max(max_temp)) %>% 
  pull(state_name)

min_temp_state <- max_min_temp %>% 
  filter(min_temp == min(min_temp)) %>% 
  pull(state_name)

print('State with the highest temperature (C) in 2021:')
print(max_temp_state)

print('State with the lowest temperature (C) in 2021:')
print(min_temp_state)

```

Great!  Now, let's take a look at average temperatures in these states over the course of the year:

```{r temp-plot, fig.width=6}
# Prepare data for plotting:
az_nd <- state_2021 %>% 
  mutate(tavg = as.numeric(tavg), # Remember to convert from character
         date = lubridate::ymd(date)) %>% 
  filter(state_name %in% c(max_temp_state, min_temp_state)) 


ggplot(az_nd, aes(x = date, y = tavg, color = state_name)) + 
  geom_line() + 
  stat_smooth() + 
  theme_bw() + 
  theme(legend.title = element_blank()) + 
  labs(x = '', y = 'Average Temperature (C)')
```



# County Data

Next, let's take a look at county data.  Specifically, let's pull county data for all counties and identify which counties in the US have the most extreme temperature swings: the largest difference between min and max temperatures on any given day over the past ~5 years.

First, let's load the county data:
```{r load-county}
county_2021 <-  read_nclimgrid_epinoaa(
  beginning_date = '2017-01-01', 
  end_date = '2021-12-31', 
  spatial_res = 'cty')

```

We are starting to get some larger data in.  The county dataset we just loaded is getting up there in MB:
```{r object_size}
as.numeric(object.size(county_2021)) / 2^20
```

Next, let's take a shot at calculating the temperature extremes for every day and every county:
```{r temp-extremes}
temp_differences <- county_2021 %>%
  mutate(tmax = as.numeric(tmax),
         tmin = as.numeric(tmin),
         temp_difference = tmax - tmin)
```

Now, we'll find the top ten counties for each year that had the highest daily temperature swing in that year:
```{r temp-year}
rank_difference <- temp_differences %>% 
  select(YEAR, state_name, region_name, temp_difference) %>% 
  group_by(YEAR) %>% 
  arrange(temp_difference) %>% 
  top_n(10) %>% 
  mutate(rank = dense_rank(desc(temp_difference))) %>% 
  arrange(rank) 

head(rank_difference)
```

Looks like counties in Kansas have a fair amount of temperature swings.  Let's take another look at these data: 
```{r rank-year-plot, fig.width=6, fig.height=5}
library(ggrepel)

county_summ <- rank_difference %>% 
  ungroup() %>% 
  group_by(region_name, state_name) %>% 
  summarize(count = n(),
            avg_temp_diff = mean(temp_difference))

ggplot(county_summ, 
       aes(x = avg_temp_diff, 
           y = count, 
           label = region_name)
       ) + 
  geom_text_repel() + 
  geom_point() + 
  theme_bw() + 
  labs(x = 'Avg Annual Maximum Daily Temperature Difference (2017-2021)',
       y = 'Number of Times Ranked \nin Top 10 Counties over 5yrs')
```

Wow!  Looks like Greeley County, KS takes the prize! It ranked in the top 10 counties for greatest difference between daily minimum and maximum temperatures every year for the past 5 years.  It also has the highest daily temperature difference as measured by the most extreme annual day averaged over the past 5 years.  

Let's look at states with counties in the top ten annually:
```{r state-county,  fig.width=6}
state_ranks <- rank_difference %>% 
  ungroup() %>% 
  group_by(YEAR, state_name) %>% 
  count() %>% 
  filter(n > 1)

ggplot(state_ranks, aes(x = as.numeric(YEAR), y = n, linetype = state_name)) + 
  geom_point() + 
  geom_line() +
  theme_bw() + 
  scale_y_continuous(breaks = seq(from = 1, to = 10, by = 2)) + 
  labs(x = 'Year', y = 'Number of Counties in Top 10') + 
  theme(legend.position = c(0.2, 0.3),
        legend.title = element_blank(),
        legend.background = element_rect(color = 'grey30'))


```


...And Colorado and Kansas it is!  Hopefully these examples help get you started using the nClimGrid daily data.  Please reach out to the NOAA BDP program (noaa.bdp@noaa.gov) with questions, comments, and feedback.  

## Time Series Data

As additional examples, let's practice reading in a complete historical time series for two states and two counties.

__State__

Let's take a look at a smoothed timeseries of maximum daily temperatures in NC and SC from 1951 to present. 

```{r state-tsh}
nc_sc <-  read_nclimgrid_epinoaa(
  spatial_res = 'ste',
  states = c('NC', 'SC'))

nc_sc %>% 
  mutate(tmax = na_if(as.numeric(tmax), -999.99)) %>% 
  ggplot(aes(x = date, y = tmax, color = postal_code)) +
  geom_smooth() + 
  theme_bw(12) +
  labs(x = '', y = 'Maximum Temperature')

```


__County__

Let's take a look at a time series of counties in Florida...

```{r county-tsh}
fl_county <-  read_nclimgrid_epinoaa(
  spatial_res = 'cty',
  states = c('FL'),
  counties = c('FL: Miami-Dade County', 'FL: Monroe County'))

fl_county %>% 
  mutate(tmax = na_if(as.numeric(tmax), -999.99)) %>% 
  ggplot(aes(x = date, y = tmax, color = region_name)) +
  geom_line() + 
  facet_grid( ~ region_name) + 
  theme_bw(12) +
  theme(legend.position = 'none') + 
  labs(x = '', y = 'Maximum Temperature')
```


And if we smooth the time series...

```{r cty-ts-smooth}
fl_county %>% 
  mutate(tmax = na_if(as.numeric(tmax), -999.99)) %>% 
  ggplot(aes(x = date, y = tmax, color = region_name)) +
  geom_smooth() + 
  theme_bw(12) +
  labs(x = '', y = 'Maximum Temperature')

```

## Census Tracts

Now let's practice looking at census tract information.  Let's pull census tracts from the state of Florida for 2022...

```{r cen-pull}
census <- read_nclimgrid_epinoaa(
  beginning_date = '2022-01-01',
  end_date = '2022-12-31',
  spatial_res = 'cen',
  states = c('FL')
)
```

Let's identify the census tracts that had the largest daily temperature swings over the past year...

```{r cen-temp-extremes}
cen_temp_differences <- census %>%
  mutate(tmax = as.numeric(tmax),
         tmin = as.numeric(tmin),
         temp_difference = tmax - tmin)

rank_difference <- cen_temp_differences %>% 
  select(fips, state_name, region_name, temp_difference) %>% 
  arrange(temp_difference) %>% 
  top_n(10) %>% 
  mutate(rank = dense_rank(desc(temp_difference))) %>% 
  arrange(rank) 

head(rank_difference)
```

